---
title: About the Data
---


This page provides a detailed description of the datasets used in this study, along with all preprocessing and data cleaning steps. These procedures ensure transparency and reproducibility of the analysis presented in the main report.

---

### (1) Prevalence of Alzheimer’s Dementia Data (Dhana et al., 2023)

- **Description:**  
This dataset originates from the Chicago Health and Aging Project (CHAP), a prospective population-based cohort study designed to evaluate risk factors for Alzheimer’s dementia (AD). Prevalence estimates were derived using neurocognitive assessments and demographic data collected from 10,802 participants through structured in-home interviews conducted in 3-year cycles.

  - **Neurocognitive tests:**  
Two episodic memory tests, one perceptual speed test, and the Mini-Mental State Examination (MMSE) assessing general orientation and global cognition.

  - **Demographic data:**  
Based on NHCS 2020 bridged-race population estimates.

- **Raw data acquisition:**  
The dataset was obtained by downloading the supplementary CSV file (NIHMS1888597-supplement-supinfo3.csv).

- **Data cleaning:**  
The dataset was processed by separating combined geographic identifiers into county and state variables and extracting numerical values for prevalence and confidence intervals from a single text column using regular expressions. Specifically, the regex pattern `([0-9.]+) \\(([0-9.]+), ([0-9.]+)\\)` was used to extract prevalence, lower CI, and upper CI values. Geographic names were standardized by removing suffixes (e.g., “County”) and converting to lowercase to ensure consistent merging with other datasets.

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

## Data Cleaning of Prevalence of Alzheimer’s Dementia Data 
ad_prevalence <- read.csv("NIHMS1888597-supplement-supinfo3.csv")

# Data Cleaning
ad_prevalence <- ad_prevalence %>%
  # Split county/state 
  separate(Counties, into = c("county", "state"), sep = ", ") %>%
  
  # Extract prevalence + CI using regex
  extract(
    'Prevalence',
    into = c("prevalence", "ci_lower", "ci_upper"),
    regex = "([0-9.]+) \\(([0-9.]+), ([0-9.]+)\\)",
    convert = TRUE
  ) %>%
  
  # Clean names for mapping
  mutate(
    county = str_remove(county, " County"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  )
```

### (2) Alzheimer’s Disease and Healthy Aging Data (BRFSS, 2015–2022)

-   **Description:** This dataset, compiled from the Behavioral Risk Factor Surveillance System (BRFSS), contains state-level survey responses related to health behaviors and conditions among adults aged 50 years and older. It includes seven modules covering health status, lifestyle factors, and cognitive decline.

-   **Cognitive Decline Module**: The analysis focused on four questions across four topics related to subjective cognitive decline (SCD), used as an indicator for early cognitive impairment.

-   **Raw data acquisition:** Data were retrieved programmatically via the CDC API using paginated requests with limit 5000 and offset 0, combining multiple chunks into a single dataset and saving locally as a CSV file.\ 
[View CDC Alzheimer’s Dataset Description](https://data.cdc.gov/Healthy-Aging/Alzheimer-s-Disease-and-Healthy-Aging-Data/hfr9-rurv/about_data).

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

## Data Acuisition of Alzheimer’s Disease and Healthy Aging Data
library(httr)
library(readr)
library(dplyr)

base_url <- "https://data.cdc.gov/resource/hfr9-rurv.csv"

limit <- 50000
offset <- 0

all_data <- list()

repeat {
  offset_str <- format(offset, scientific = FALSE)   # should be integer formatting (NO scientific notation)
  
  url <- paste0(base_url, "?$limit=", limit, "&$offset=", offset_str)
  cat("Fetching:", url, "\n")
  
  res <- GET(url)
  
  if (status_code(res) != 200) {
    stop("Request failed with status: ", status_code(res))
  }
  
  chunk <- read_csv(content(res, "raw"), show_col_types = FALSE)
  
  if (nrow(chunk) == 0) break
  
  all_data[[length(all_data) + 1]] <- chunk
  
  offset <- as.integer(offset + limit)
}

cdc_data <- bind_rows(all_data)

write_csv(cdc_data, "cdc_dataset.csv")
```

-   **Data cleaning:** Approximately 30% of observations contained missing values, which were retained because they reflect survey design limitations and statistical reliability rather than random absence. A summary table of missing values was generated to assess data completeness (see the table 1 in the Result page). In addition, overall category which was used in this study has only 0.07% of missingness indicating the high completness of the dataset (See table 4 in the Result page).

### (3) Demographic and Economic Data (U.S. Census Bureau)

-   **Description:** The U.S. Census Bureau provides detailed demographic and housing data through the Decennial Census and American Community Survey (ACS), including population structure, household composition, and socioeconomic indicators.

-   **Datasets used:**

    -   Total population (Decennial Census 2020): DECENNIALDHC2020.P1

    -   Population 65 Years and Over (ACS 5-year estimates 2016-2020): ACSST5Y2020.S0103

    -   SNAP (Food Stamps) participation (ACS 5-year estimates 2016-2020): ACSST5Y2020.S2201

-   **Raw data acquisition:** Datasets were downloaded as CSV files from the Census data portal after applying filters for geography (county level), year (2020), and relevant topics.\
[Access Census Data](https://data.census.gov/table?t=Populations+and+People&g=010XX00US$0500000&y=2020&d=DEC+Demographic+and+Housing+Characteristics)

-   **Data cleaning:** Raw Census datasets with multi-row headers and complex column structures were standardized by restructuring column names, selecting relevant variables, and converting population counts to numeric format. Geographic identifiers were harmonized by splitting county and state fields and standardizing text formatting, enabling integration with prevalence and survey datasets. Other datasets from U.S. Census Bureau were processed with the same procedure.

```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

# DATA CLEANING: Senior Population & Percentages
pop_raw <- read_csv("seniorpop5Y2020.csv", show_col_types = FALSE)

pop_clean <- pop_raw %>%
  slice(-1) %>%                      # Clean multi-row headers
  select(
    location = NAME,                 # selecting relevant variables
    total_pop = S0103_C01_001E,
    senior_pop = S0103_C02_001E
  ) %>%
  mutate(
    total_pop = as.numeric(total_pop),# converting population counts to numeric format
    senior_pop = as.numeric(senior_pop),
    pct_senior = (senior_pop / total_pop) * 100,
    location = str_trim(location)
  ) %>%
  drop_na(total_pop, senior_pop) %>%
  separate(location, into = c("county", "state"), sep = ", ") %>% #splitting county and state fields
  mutate(
    county = str_remove(county, " County| Parish| Borough| Census Area| Municipality"), #text formatting
    county_lower = tolower(county), # for integration
    state_lower = tolower(state)
  ) %>%
  filter(state_lower != "district of columbia")        
```

### (4) Geographic Data (U.S. Department of Transportation National Transportation Atlas Database (NTAD))

- **Description:**  
Geographic boundary data were used to calculate population density for both total and senior populations at the county level. County-level area information was obtained from the U.S. Department of Transportation National Transportation Atlas Database (NTAD), which provides spatial and geographic attributes for U.S. counties. Two area-related variables are included in the dataset: land area (`ALAND`) and water area (`AWATER`), both measured in square meters. In this study, only land area (`ALAND`) was used to calculate population density, as it more accurately reflects habitable space.

- **Raw data acquisition:**  
The county-level area dataset (NTAD_Counties.csv) was downloaded from the Department of Transportation’s NTAD database. Population data were obtained separately from the U.S. Census Bureau (Decennial Census 2020).

- **Data cleaning:**  
Population and geographic datasets were cleaned and standardized using county-level geographic identifiers (GEOID). In the population dataset, relevant variables were selected, geographic identifiers were reformatted, and population counts were converted to numeric values. Non-county entries were removed to ensure consistency. In the NTAD dataset, land area (`ALAND`) was extracted and matched to county GEOIDs. These cleaned datasets were subsequently merged to calculate population density as population divided by land area.


[Access Counties Data](https://geodata.bts.gov/datasets/usdot::counties/explore?location=28.418990%2C0.314282%2C1&showTable=true)\
[View Counties Dataset Description](https://catalog.data.gov/dataset/counties2)


```{r, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}

# POPULATION DATA 

pop_raw <- read_csv("totalpop2020.csv", show_col_types = FALSE)

pop_clean <- pop_raw %>%
  select(GEO_ID, NAME, `P1_001N`) %>%          # use GEOID
  rename(
    geoid = GEO_ID,
    location = NAME,
    population = `P1_001N`
  ) %>%
  filter(str_detect(location, ",")) %>%
  mutate(
    geoid = str_remove(geoid, "0500000US"),   # Non-county entries were removed
    population = as.numeric(population)       # converted to numeric values
  ) %>%
  drop_na(population)


# 2. NTAD AREA DATA (use GEOID)

ntad_raw <- read_csv("NTAD_Counties.csv", show_col_types = FALSE)

ntad_clean <- ntad_raw %>%
  mutate(
    geoid = as.character(GEOID),               # use GEOID
    area_m2 = ALAND
  ) %>%
  select(geoid, area_m2)
```