---
title: Results
format: html
editor: visual
---

## 1. Exploratory Data Analysis

```{r, include = FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(maps)
library(patchwork)
library(plotly)
library(gt)
library(kableExtra)
library(scales)
library(lubridate)


ad_data <- read_csv("cdc_dataset.csv")

#glimpse(ad_data)
#summary(ad_data)
```

### 1.1. Variable-Level Missing Value Analysis

To assess data completeness in the Alzheimer’s Disease and Healthy Aging dataset (AD dataset), a summary table of missing values was generated for each variable (Table 1). Missing values (NAs) in data_value and confidence interval variables primarily reflect statistical or sampling limitations rather than data errors. Detailed explanations are provided in the data_value_footnote_symbol and data_value_footnote fields.

The first level of stratification contains no missing values, whereas most missingness occurs in the second level, particularly for observations labeled “Overall” within age groups. These missing values are intentionally introduced to avoid duplication across stratification levels. Missing geolocation values correspond to the “United States” category, where geographic coordinates are not required. Overall, the dataset remains suitable for analysis.

Data integrity levels were defined as follows:

Missing (%) = 0 → Complete; Missing (%) \< 20 → High Integrity; Missing (%) \< 50 → Conditional;

```{r, echo = FALSE}
# 1. Enhanced Missing Value Summary Table
na_summary_table <- ad_data %>%
  summarise(across(everything(), list(
    count = ~sum(is.na(.)),
    pct = ~round(mean(is.na(.)) * 100, 2) #%
  ))) %>%
  pivot_longer(
    everything(), 
    names_to = c("Variable Name", ".value"), 
    names_sep = "_(?!.*_)" # Regex to split only at the last underscore
  ) %>%
  mutate(Status = case_when(
    pct == 0 ~ "Complete",
    pct < 20 ~ "High Integrity",
    pct < 50 ~ "Conditional",
    TRUE ~ "Limited"
  )) %>%
  rename(`NA Count` = count, `Missing (%)` = pct)

na_summary_table %>%
  kbl(caption = "Table 1: Variable-Level Completeness Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  column_spec(3, bold = TRUE, color = spec_color(na_summary_table$`Missing (%)`, end = 0.7))
```

### 1.2. Data Integrity and Module Summary

The AD dataset consists of seven modules containing survey questions related to Alzheimer’s disease and its prevention. For each module, the number of observations and the proportion of missing values were calculated (Table 2).

Responses are stratified first by age group and then by gender and race/ethnicity. Observations are reported at the state level for two age groups (50–64 and 65+), with the “Overall” category representing the combined population.

```{r, echo = FALSE}

# 1. Missing Value Summary
na_summary <- ad_data %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  tidyr::pivot_longer(everything(), names_to = "Variable Name", values_to = "NA Count")

# 2. Filtering for Cognitive Decline and Standardizing Geography
scd_data <- ad_data %>%
  filter(class == "Cognitive Decline") %>% #class=Module
  rename(state = locationdesc)

# 3. Generating the Publication Table for the Result Section
# Using dplyr::group_by to be explicit and avoid conflicts
summary_table <- ad_data %>%
  dplyr::group_by(Module = class) %>%
  dplyr::summarise(
    Observations = n(),
    `Missing Values (%)` = round(mean(is.na(data_value)) * 100, 1),
    .groups = 'drop'
  ) %>%
  dplyr::mutate(
    Status = case_when(
      `Missing Values (%)` < 20 ~ "Validated",
      `Missing Values (%)` < 40 ~ "Retained",
      TRUE ~ "Filtered"
    )
  )

# Render with kableExtra (explicitly calling the package)
summary_table %>%
  kableExtra::kbl(caption = "Table 2: Data Integrity and Module Summary") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```

### 1.3. Survey questions in Cognitive Decline Module

This study focuses on the Cognitive Decline module, which includes measures of subjective cognitive decline (SCD), a potential early indicator of cognitive impairment.

The module contains four key questions:

Q30: Frequency of subjective cognitive decline Q31: Impact on daily activities Q41: Need for assistance Q42: Experience seeking medical help

A detailed mapping of these questions is provided in Table 3.

```{r, echo = FALSE}
# Select Cognitive Decline Module
scd_data <- ad_data %>%
  filter(class == "Cognitive Decline") %>% 
  rename(state = locationdesc)

# Survey Question Mapping Table
summary_lookup <- tibble::tribble(
  ~questionid, ~summary,
  "Q30", "Subjective cognitive decline (SCD) freq",
  "Q31", "Daily activity impact",
  "Q41", "Need assistance",
  "Q42", "Discussed with doctor"
)

question_map <- scd_data %>%
  select(questionid, topic, question) %>%
  distinct() %>%
  left_join(summary_lookup, by = "questionid") %>%
  arrange(questionid, topic) %>%
  select(questionid, summary, topic, question) # Reordering columns for clarity

question_map %>%
  kbl(caption = "Table 3: Mapping of Question ID to Topic and Survey Question",
      col.names = c("Question ID", "Summary", "Topic", "Survey Question")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = T) %>%
  column_spec(1, width = "100px") %>%
  column_spec(2, width = "170px") %>%
  column_spec(3, width = "250px") %>%
  column_spec(4, width = "350px") %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2")
```

### 1.4. Data Integrity of survey questions in Cognitive Decline Module

The proportion of missing data across the four Cognitive Decline questions ranges from approximately 33% to 38% (Figure 1). Q30 exhibits the lowest missingness (33%), while Q41 shows the highest (38%), with Q31 and Q42 at similar intermediate levels.

This relatively consistent level of missingness across questions suggests a systematic pattern likely driven by survey design and stratification structure rather than random data loss. Despite moderate missingness, the data remain comparable across questions and are appropriate for descriptive and comparative analysis.

Importantly, when restricting to overall stratifications, the proportion of missing values in data_value is extremely low (\~0.07%), indicating high data completeness and supporting the reliability of analyses based on aggregated groups (Table 4).

```{r, echo=FALSE}
## Missing data% by survey questions
# Theme of choice
theme_modern <- theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(hjust = 1),
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(size = 12, hjust = 0, color = "grey40"),
    legend.position = "top",
 #   plot.margin = margin(10, 20, 10, 10)
  )
modern_blue <- c("#1D3557", "#2C7FB8", "#41B6C4", "#7FCDBB", "#A6BDD9")

na_by_question <- scd_data %>%
  group_by(questionid) %>%
  summarise(na_rate = mean(is.na(data_value)), .groups = "drop") 

ggplot(na_by_question,
       aes(x = reorder(questionid, desc(questionid)),
           y = na_rate)) +
  geom_col(fill = "#2C7FB8", width = 0.6) +   # thinner bars
  coord_flip() +
  scale_y_continuous(limits = c(0, 1)) +      # proportion scale
  labs(
    title = "Proportion of Missing Data by Survey Question",
    x = "Question ID",
    y = "Proportion of Missing Values",
    caption = "Figure 1: Proportion of Missing Data by Survey Question in Cognitive Decline Module"
  ) +
  geom_text(aes(label = scales::percent(na_rate, accuracy = 1)),
          hjust = -0.1, size = 3.5)+
  theme_modern


## Clean aggregated Observations only
# Overall strata only
overall_all <- ad_data %>%
  filter(
    stratification1 == "Overall",
    stratificationcategoryid2 == "OVERALL"
  )

# Compute NA %
na_all <- overall_all %>%
  summarise(
    total_n = n(),
    na_n = sum(is.na(data_value)),
    na_percent = mean(is.na(data_value)) * 100
  )

# Cognitive Decline only
overall_cd <- ad_data %>%
  filter(
    class == "Cognitive Decline",
    stratification1 == "Overall",
    stratificationcategoryid2 == "OVERALL"
  )

na_cd <- overall_cd %>%
  summarise(
    total_n = n(),
    na_n = sum(is.na(data_value)),
    na_percent = mean(is.na(data_value)) * 100
  )


# Combine results
na_summary_table <- bind_rows(
  na_all %>% mutate(Group = "All Classes"),
  na_cd %>% mutate(Group = "Cognitive Decline")
) %>%
  select(
    Group,
    Observations = total_n,
    `Missing Values (n)` = na_n,
    `Missing Values (%)` = na_percent
  ) %>%
  mutate(
    `Missing Values (%)` = round(`Missing Values (%)`, 2),
    Status = case_when(
      `Missing Values (%)` == 0 ~ "Complete",
      `Missing Values (%)` < 1 ~ "Highly Complete",
      `Missing Values (%)` < 20 ~ "Validated",
      TRUE ~ "Check Needed"
    )
  )

# Render table
na_summary_table %>%
  kableExtra::kbl(
    caption = "Table 4: Missingness in Overall Stratified Data",
    align = "lcccc"
  ) %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  kableExtra::column_spec(1, bold = TRUE) %>%
  kableExtra::column_spec(4, bold = TRUE, color = "white", background = "#2C7FB8") %>%
  kableExtra::add_header_above(c(" " = 1, "Sample Size" = 1, "Missingness" = 3))

```

## 2. State-Level Analysis

### 2.1. SCD Comparison of High AD Prevalence States to the U.S. National Level (2015–2022)

To assess whether states with the highest Alzheimer’s dementia (AD) prevalence—Maryland (12.9%), New York (12.7%), Mississippi (12.5%), and Florida (12.5%)—share similar patterns in subjective cognitive decline (SCD), response frequencies from the Cognitive Decline module were analyzed. Stratification by gender and race/ethnicity was excluded, and only aggregated state-level data from 2015 to 2022 were used.

In the jitter plot (Figure 2), these four states were highlighted against the distribution of all states (gray), with the U.S. national aggregate shown in black as a reference. Maryland consistently exhibited lower SCD-related response frequencies than the national average, whereas Mississippi showed higher values across multiple indicators. This suggests that older adults in Mississippi may report greater awareness of cognitive decline and a higher need for assistance, aligning with its elevated AD prevalence. In contrast, Maryland demonstrated an inverse pattern, with lower SCD reporting despite high AD prevalence. Florida and New York showed values similar to the national distribution, indicating no distinct deviation in SCD-related measures.

```{r, fig.width=9, fig.height=5, out.width="100%", echo=FALSE, warning=FALSE, message=FALSE}

library(dplyr)
library(ggplot2)
library(plotly)

# 1. Prepare Data
summary_lookup <- tibble::tribble(
  ~questionid, ~summary,
  "Q30", "Subject Cognitive Decline freq",
  "Q31", "Daily activity impact",
  "Q41", "Need assistance",
  "Q42", "Discussed with doctor"
)

plot_data <- scd_data %>%
  filter(
    stratification1 == "Overall",
    stratificationcategoryid2 == "OVERALL",
    yearstart == yearend,       
    !is.na(data_value)
  ) %>%
  left_join(summary_lookup, by = "questionid") %>%
  mutate(
    summary = factor(summary, levels = c(
      "Subject Cognitive Decline freq", 
      "Daily activity impact",          
      "Need assistance",                
      "Discussed with doctor"           
    )),
    highlight_group = case_when(
      state == "Maryland" ~ "Maryland",
      state == "New York" ~ "New York",
      state == "Mississippi" ~ "Mississippi",
      state == "Florida" ~ "Florida",
      state %in% c("United States, DC & Territories") ~ "US National",
      TRUE ~ "Other States"
    ),
    highlight_group = factor(highlight_group, 
                             levels = c("Maryland", "New York", "Mississippi", 
                                        "Florida", "US National", "Other States"))
  )

pub_palette <- c(
  "Other States" = "#e0e0e0",     
  "US National" = "#000000",      
  "Florida" = "#d55e00",          
  "Mississippi" = "#009e73",      
  "New York" = "#56b4e9",         
  "Maryland" = "#0072b2"          
)

# 2. Build the Plot (No Caption)
p <- ggplot(plot_data, aes(x = data_value, y = summary, 
                           text = paste0("State: ", state, "\n",
                                         "Percentage: ", data_value, "%\n",
                                         "Year: ", yearend))) +
  geom_point(data = filter(plot_data, highlight_group == "Other States"),
             color = "gray80", size = 2, alpha = 0.6, 
             position = position_jitter(height = 0.15, width = 0)) +
  geom_point(data = filter(plot_data, highlight_group != "Other States"),
             aes(color = highlight_group), 
             size = 4, alpha = 0.9,
             position = position_jitter(height = 0.05, width = 0)) + 
  scale_color_manual(values = pub_palette) +
  scale_y_discrete(limits = rev) +
  labs(
    title = "Subjective Cognitive Decline Survey Responses",
    subtitle = "Highlighting states with high modeled Alzheimer's prevalence vs. the National Average",
    x = "Percentage of Respondents (%)",
    y = NULL,
    color = "Region"
  ) +
  theme_minimal(base_size = 12) + # Slightly smaller base font so things fit better
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", margin = margin(b = 10)),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.5),
    panel.grid.minor = element_blank()
  )

# 3. Interactive Convert (Strictly inheriting the chunk size)
interactive_plot <- ggplotly(p, tooltip = "text") %>%
  layout(
    autosize = TRUE,
    # Keep the legend horizontal, tightly packed under the x-axis
    legend = list(orientation = "h", x = 0.5, xanchor = "center", y = -0.25),
    # Tighter margins so the graph doesn't artificially expand
    margin = list(t = 40, b = 80, l = 10, r = 10)
  ) %>%
  config(displaylogo = FALSE)

interactive_plot
```

### 2.2. Longitudinal SCD Comparison (2015–2022)

To evaluate temporal trends, SCD-related response frequencies were examined longitudinally (Figure 3). The overall patterns remained consistent with cross-sectional findings. Maryland showed persistently lower levels of reported cognitive decline, functional difficulty, and need for assistance compared to the national average across the study period. Conversely, Mississippi consistently exhibited higher levels across these measures, suggesting a sustained burden of perceived cognitive decline.

No substantial differences were observed among the four states in terms of discussions with healthcare providers. Overall, no consistent pattern in SCD measures was identified that could explain the shared high AD prevalence across these states.

```{r, fig.height=7, fig.width=10, echo=FALSE, message=FALSE, warning=FALSE}

#  Prepare the Data
plot_trend_data <- scd_data %>%
  filter(
    stratification1 == "Overall",
    stratificationid2 == "OVERALL",
    yearstart == yearend,       
    !is.na(data_value)          
  ) %>%
  left_join(summary_lookup, by = "questionid") %>%
  mutate(
    summary = factor(summary, levels = c(
      "Subject Cognitive Decline freq", # Q30
      "Daily activity impact",          # Q31
      "Need assistance",                # Q41
      "Discussed with doctor"           # Q42
    )),
    
    highlight_group = case_when(
      state == "Maryland" ~ "Maryland",
      state == "New York" ~ "New York",
      state == "Mississippi" ~ "Mississippi",
      state == "Florida" ~ "Florida",
      state %in% c("United States, DC & Territories") ~ "US National",
      TRUE ~ "Other States"
    ),
    highlight_group = factor(highlight_group, 
                             levels = c("Maryland", "New York", "Mississippi", 
                                        "Florida", "US National", "Other States"))
  )

# 3. Define the Publication Palette
pub_palette <- c(
  "Other States" = "#e0e0e0",     
  "US National"  = "#000000",      
  "Florida"      = "#d55e00",          
  "Mississippi"  = "#009e73",      
  "New York"     = "#56b4e9",         
  "Maryland"     = "#0072b2"          
)

# 4. Build the Plot
ggplot(plot_trend_data, aes(x = yearend, y = data_value, group = state)) +
  
  geom_line(data = filter(plot_trend_data, highlight_group == "Other States"),
            color = "#e0e0e0", linewidth = 0.5, alpha = 0.5) +
  
  geom_line(data = filter(plot_trend_data, highlight_group != "Other States"),
            aes(color = highlight_group), linewidth = 0.9) +
  
  geom_point(data = filter(plot_trend_data, highlight_group != "Other States"),
             aes(color = highlight_group), size = 1.8) +
  
  scale_color_manual(values = pub_palette) +
  scale_x_continuous(breaks = seq(2015, 2022, by = 1)) +
  
  facet_wrap(~summary, scales = "free_y", ncol = 2) +
  
  labs(
    title = "Longitudinal Trends in Subjective Cognitive Decline (2015–2022)",
    subtitle = "Comparing high AD-prevalence states against the national distribution",
    x = "Survey Year",
    y = "Percentage of Respondents (%)",
    color = "Region",
    caption = "Figure 3: Subjective Cognitive Decline Survey Response Changes over Time (2015-2022)"
  ) +
  
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(color = "gray40", margin = margin(b = 15)),
    strip.text = element_text(face = "bold", size = 12),
    strip.background = element_rect(fill = "gray95", color = NA),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "gray95", linewidth = 0.5),
    legend.position = "bottom",
    plot.margin = margin(t = 20, r = 20, b = 20, l = 10)
  ) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE, 
                              override.aes = list(linewidth = 2)))
```

### 2.3. Covid19 Motality Comparision

To estimate the condition of health environment of the high AD prevalence states (highlighted by \*), I mapped the COVID19 mortality rates on the national map. Although the causes of high mortality late will vary such as vaccination uptake, public health policies, the demorgraphuc and heath factors such as obeity rates and age, we cannot deny it can be a good indicator total health care environment. However, the four states COVID19 death rates were very representing their different healthcare environment.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load data
covid <- read.csv("covid_processed.csv")

# Prepare data
covid <- covid %>%
  mutate(
    death_rate = deaths / pop * 100000,
    date_object = ymd(date)
  )

# Filter 2020
covid_2020 <- covid %>%
  filter(year(date_object) == 2020)

# Summarize by state
covid_state <- covid_2020 %>%
  group_by(state_name) %>%
  summarize(
    death_rate = mean(death_rate, na.rm = TRUE),
    deaths = sum(deaths, na.rm = TRUE),
    pop = first(pop)
  ) %>%
  mutate(state_lower = tolower(state_name))

# Map data
us_map <- map_data("state")

map_data_covid <- left_join(us_map, covid_state,
                            by = c("region" = "state_lower"))

# States to highlight
highlight_states <- c("maryland", "new york", "mississippi", "florida")

# state.name and state.center are pre-loaded in R
star_data <- data.frame(
  state = tolower(state.name),
  long = state.center$x,
  lat = state.center$y
) %>%
  filter(state %in% c("maryland", "new york", "mississippi", "florida"))


# Create ggplot with hover text
p <- ggplot(map_data_covid, aes(long, lat, group = group)) +
  geom_polygon(
    aes(fill = death_rate,
        text = paste0(
          "State: ", state_name,
          "<br>Death Rate: ", round(death_rate, 2),
          "<br>Total Deaths: ", deaths,
          "<br>Population: ", pop
        )),
    color = "white"
  ) +
  
  geom_text(data = star_data,
            aes(long, lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +
  
  scale_fill_gradient(
    low = "lightyellow",
    high = "#800026",
    name = "Deaths per 100K",
    na.value = "grey90"
  ) +
  
  coord_fixed(1.3) +
  labs(title = "Figure 4: US COVID-19 Death Rates (2020)") +
  theme_void()

# Convert to interactive plot
ggplotly(p, tooltip = "text")
```

## 3. County-Level Analysis of AD Prevalence

The absence of consistent state-level patterns suggests that aggregated statistics may obscure important regional variation. Therefore, county-level AD prevalence data from Dhana et al. (2023) were visualized to explore geographic patterns (Figure 4).

Two prominent curved geographic patterns were observed: one extending along the eastern-to-southern boundary and another spanning north-to-south across the central United States. These patterns suggest the presence of region-specific determinants not captured at the state level.

```{r, echo=FALSE}

ad_prevalence <- read.csv("NIHMS1888597-supplement-supinfo3.csv")

# Data Cleaning
ad_prevalence <- ad_prevalence %>%
  # Split county/state 
  separate(Counties, into = c("county", "state"), sep = ", ") %>%
  
  # Extract prevalence + CI using regex
  extract(
    'Prevalence',
    into = c("prevalence", "ci_lower", "ci_upper"),
    regex = "([0-9.]+) \\(([0-9.]+), ([0-9.]+)\\)",
    convert = TRUE
  ) %>%
  
  # Clean names for mapping
  mutate(
    county = str_remove(county, " County"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  )
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
county_map <- map_data("county")

county_map_data <- county_map %>%
  left_join(ad_prevalence,
            by = c("subregion" = "county_lower",
                   "region" = "state_lower"))

# 3. Create the Star Data using built-in R state centers
# state.name and state.center are pre-loaded in R
star_data <- data.frame(
  state = tolower(state.name),
  long = state.center$x,
  lat = state.center$y
) %>%
  filter(state %in% c("maryland", "new york", "mississippi", "florida"))

# Plot
ggplot(county_map_data, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = prevalence), color = NA) +

  
  scale_fill_gradientn(
    colors = c("#F0FAF9", "#B2E2E2", "#66C2A4", "#2CA25F", "#006D2C"),
    name = "AD Prevalence (%)",
    na.value = "grey90"
  ) +
  
    
  geom_text(data = star_data,
            aes(x=long, y=lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +    
  
  coord_fixed(1.3) +
  labs(title = "Figure 5: Alzheimer’s Disease Prevalence (County-Level, Age ≥65)") +
  theme_void()
```

### 3.1. Population effect

#### 3.1.1. Total Population

We first hypothesized that these patterns may reflect population distribution. County-level total population data from the U.S. Census were visualized (Figure 5a). Due to skewed population distribution, a log transformation was applied. However, the resulting map did not reproduce the observed AD prevalence patterns.

Population density was then calculated using county area data, and the resulting density map (Figure 5b) showed some linear clustering of populated regions. However, these did not align with the patterns observed in AD prevalence, suggesting that total population alone does not explain the geographic distribution.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# ---------------------------------------------------------
# 1. POPULATION DATA (extract FIPS)
# ---------------------------------------------------------
pop_raw <- read_csv("totalpop2020.csv", show_col_types = FALSE)

pop_clean <- pop_raw %>%
  select(GEO_ID, NAME, `P1_001N`) %>%
  rename(
    geoid = GEO_ID,
    location = NAME,
    population = `P1_001N`
  ) %>%
  filter(str_detect(location, ",")) %>%
  mutate(
    geoid = str_remove(geoid, "0500000US"),
    population = as.numeric(population)
  ) %>%
  drop_na(population)

# ---------------------------------------------------------
# 2. NTAD AREA DATA (use GEOID)
# ---------------------------------------------------------
ntad_raw <- read_csv("NTAD_Counties.csv", show_col_types = FALSE)

ntad_clean <- ntad_raw %>%
  mutate(
    geoid = as.character(GEOID),
    area_m2 = ALAND
  ) %>%
  select(geoid, area_m2)

# ---------------------------------------------------------
# 3. MERGE + DENSITY (NO DUPLICATES)
# ---------------------------------------------------------
pop_density <- pop_clean %>%
  left_join(ntad_clean, by = "geoid") %>%
  mutate(
    density_km2 = population / (area_m2 / 1e6)
  )

# ---------------------------------------------------------
# 4. MAP DATA (match FIPS)
# ---------------------------------------------------------
us_counties <- map_data("county")

# maps package does NOT have FIPS → need alternative matching
# fallback: still use name matching but CLEANED

pop_density_named <- pop_density %>%
  separate(location, into = c("county", "state"), sep = ", ") %>%
  mutate(
    county = str_remove(county, " County| Parish| Borough| Census Area"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  )

map_data_all <- us_counties %>%
  left_join(pop_density_named,
            by = c("subregion" = "county_lower",
                   "region" = "state_lower"))

# Remove DC
map_data_clean <- map_data_all %>%
  filter(region != "district of columbia")

# Cap density
upper_limit <- quantile(map_data_clean$density_km2, 0.99, na.rm = TRUE)

# ---------------------------------------------------------
# 5. STAR LOCATIONS
# ---------------------------------------------------------
star_locs <- data.frame(
  long = c(-76.64, -75.52, -89.39, -81.51),
  lat  = c(39.04, 43.29, 32.35, 27.66)
)

# ---------------------------------------------------------
# 6. THEME
# ---------------------------------------------------------
theme_map <- theme_void() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
  )

# ---------------------------------------------------------
# 7. MAP 1: POPULATION
# ---------------------------------------------------------
p1 <- ggplot(map_data_clean, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = population), color = NA) +
  scale_fill_gradient(
    trans = "log10",
    low = "#f2e5ff",
    high = "#4b0082",
    name = "Population (log10)",
    na.value = "grey90"
  ) +
  geom_text(data = star_locs,
            aes(x = long, y = lat, label = "*"),
            color = "yellow", size = 5,
            inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 6a: US Total Population (2020)") +
  theme_void()

p1

# ---------------------------------------------------------
# 8. MAP 2: DENSITY
# ---------------------------------------------------------
p2 <- ggplot(map_data_clean, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pmin(density_km2, upper_limit)), color = NA) +
  scale_fill_gradient(
    low = "#f2e5ff",
    high = "#4b0082",
    name = "Density (per km²)",
    labels = comma,
    na.value = "grey90"
  ) +
  geom_text(data = star_locs,
            aes(x = long, y = lat, label = "*"),
            color = "yellow", size = 5,
            inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 6b: US Total Population Density (2020)") +
  theme_void()

p2
```

#### 3.1.2. Senior Population (Age ≥65)

Option2 Given that AD prevalence is defined among individuals aged 65 years and older, senior population data were examined. Both the density (Figure 7b) and proportion (Figure 7c) of the senior population were visualized. While these measures provided some regional variation, data coverage was limited, and extreme values in certain counties reduced interpretability. Consequently, no clear alignment with AD prevalence patterns was observed, and further high-resolution data would be required for conclusive analysis.

Option1 Given that Alzheimer’s disease (AD) prevalence is defined among individuals aged 65 years and older, senior population characteristics were further examined. Both the density (Figure 7b) and proportion (Figure 7c) of the senior population were visualized to assess potential geographic patterns. However, consistent with the findings from total population analyses, no distinct or meaningful spatial patterns were observed.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# ---------------------------------------------------------
# 1. DATA CLEANING: Total Population (pop_raw)
# ---------------------------------------------------------
pop_raw <- read_csv("totalpop2020.csv", show_col_types = FALSE)

pop_clean <- pop_raw %>%
  slice(-1) %>% 
  select(
    location = NAME,
    total_pop = P1_001N
  ) %>%
  mutate(
    total_pop = as.numeric(total_pop),
    location = str_trim(location)
  ) %>%
  drop_na(total_pop) %>%
  separate(location, into = c("county", "state"), sep = ", ") %>%
  mutate(
    county = str_remove(county, " County| Parish| Borough| Census Area| Municipality"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  ) 

# ---------------------------------------------------------
# 2. DATA CLEANING: Senior Population "NIHMS1888597-supplement-supinfo3.csv"
# ---------------------------------------------------------
ad_clean <- read.csv("NIHMS1888597-supplement-supinfo3.csv")

ad_clean <- ad_clean %>%
  select(
    location = 2, 
    senior_pop_thousands = 3
  ) %>%
  mutate(
    senior_pop = as.numeric(senior_pop_thousands) * 1000,
    location = str_trim(location)
  ) %>%
  drop_na(senior_pop) %>%
  separate(location, into = c("county", "state"), sep = ", ") %>%
  mutate(
    county = str_remove(county, " County| Parish| Borough| Census Area| Municipality"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  )

# ---------------------------------------------------------
# 3. LAND AREA & CALCULATIONS
# ---------------------------------------------------------
ntad_raw <- read_csv("NTAD_Counties.csv", show_col_types = FALSE)

ntad_clean <- ntad_raw %>%
  mutate(
    county_lower = tolower(str_remove(NAME, " County| Parish| Borough")),
    area_sq_km = ALAND / 1000000 
  ) %>%
  select(county_lower, area_sq_km)

# Join datasets
final_data <- pop_clean %>%
  left_join(ad_clean, by = c("county_lower", "state_lower")) %>%
  left_join(ntad_clean, by = "county_lower") %>%
  mutate(
    pct_senior = (senior_pop / total_pop) * 100,
    senior_density = senior_pop / area_sq_km
  ) %>%
  distinct(state_lower, county_lower, .keep_all = TRUE)

# ---------------------------------------------------------
# 4. BASE MAP SETUP
# ---------------------------------------------------------
us_counties <- map_data("county") %>%
  filter(region != "district of columbia")

map_data_all <- us_counties %>%
  left_join(final_data,
            by = c("subregion" = "county_lower",
                   "region" = "state_lower"))

star_locs <- data.frame(
  state = c("Maryland", "New York", "Mississippi", "Florida"),
  long = c(-76.64, -75.52, -89.39, -81.51),
  lat = c(39.04, 43.29, 32.35, 27.66)
)

# ---------------------------------------------------------
# 5. BUILD THE THREE MAPS
# ---------------------------------------------------------

# (a) Senior Population Map (Total Count)
p1 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = senior_pop), color = NA) +
  scale_fill_gradient(
    trans = "log10", 
    low = "#f2e5ff", high = "#4b0082", 
    name = "Seniors (log)", na.value = "grey90"
  ) +
  geom_text(data = star_locs, aes(x = long, y = lat, label = "*"),
            color = "yellow", size = 7, inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 7a: Total Senior Population (65+)") +
  theme_void() + theme(legend.position = "bottom")

# (b) Senior Population Density Map (km²)
p2 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  # FIX: geom_polygon was missing here!
  geom_polygon(aes(fill = senior_density), color = NA) + 
  scale_fill_gradient(
    trans = "log10",  
    low = "#e5f5f9", high = "#006d2c", 
    name = "Seniors/km² (log)", na.value = "grey90"
  ) +
  geom_text(data = star_locs, aes(x = long, y = lat, label = "*"),
            color = "yellow", size = 7, inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 7b Senior Population Density (65+)") +
  theme_void() + theme(legend.position = "bottom")

# (c) % of Senior Population Map
p3 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pct_senior), color = NA) +
  scale_fill_gradient(
    low = "#fff7bc", high = "#d95f0e", 
    name = "% Senior", na.value = "grey90"
  ) +
  geom_text(data = star_locs, aes(x = long, y = lat, label = "*"),
            color = "yellow", size = 7, inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 7c Percentage of Senior Population (65+)") +
  theme_void() + theme(legend.position = "bottom")

p1
p2
p3

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# DATA CLEANING: Senior Population & Percentages

pop_raw <- read_csv("seniorpop5Y2020.csv", show_col_types = FALSE)

pop_clean <- pop_raw %>%
  slice(-1) %>%
  select(
    location = NAME,
    total_pop = S0103_C01_001E,
    senior_pop = S0103_C02_001E
  ) %>%
  mutate(
    total_pop = as.numeric(total_pop),
    senior_pop = as.numeric(senior_pop),
    pct_senior = (senior_pop / total_pop) * 100,
    location = str_trim(location)
  ) %>%
  drop_na(total_pop, senior_pop) %>%
  separate(location, into = c("county", "state"), sep = ", ") %>%
  mutate(
    county = str_remove(county, " County| Parish| Borough| Census Area| Municipality"),
    county_lower = tolower(county),
    state_lower = tolower(state)
  ) %>%
  filter(state_lower != "district of columbia")

# ---------------------------------------------------------
# 2. LAND AREA & OUTLIER REMOVAL
# ---------------------------------------------------------
ntad_raw <- read_csv("NTAD_Counties.csv", show_col_types = FALSE)

ntad_clean <- ntad_raw %>%
  mutate(
    county_lower = tolower(str_remove(NAME, " County| Parish| Borough")),
    area_sq_km = ALAND / 1000000 # Convert to km²
  ) %>%
  select(county_lower, area_sq_km)

senior_data_final <- pop_clean %>%
  left_join(ntad_clean, by = "county_lower") %>%
  mutate(senior_density = senior_pop / area_sq_km) %>%
  distinct(state_lower, county_lower, .keep_all = TRUE)

# --- IDENTIFY AND REMOVE THE OUTLIER ---
# Find the county with the absolute highest density
outlier_county <- senior_data_final %>%
  arrange(desc(senior_density)) %>%
  slice(1)

# Print a message to the console so you know exactly what was removed
cat("Removing outlier for better map coloring:", 
    str_to_title(outlier_county$county), "County,", 
    str_to_title(outlier_county$state), 
    "( Density:", round(outlier_county$senior_density, 0), "seniors/km² )\n")

# Filter that specific county out of the dataset
senior_data_final <- senior_data_final %>%
  filter(!(county_lower == outlier_county$county_lower & state_lower == outlier_county$state_lower))

# ---------------------------------------------------------
# 3. BASE MAP SETUP
# ---------------------------------------------------------
us_counties <- map_data("county") %>%
  filter(region != "district of columbia")

map_data_all <- us_counties %>%
  left_join(senior_data_final,
            by = c("subregion" = "county_lower",
                   "region" = "state_lower"))

star_locs <- data.frame(
  state = c("Maryland", "New York", "Mississippi", "Florida"),
  long = c(-76.64, -75.52, -89.39, -81.51),
  lat = c(39.04, 43.29, 32.35, 27.66)
)

# ---------------------------------------------------------
# 4. BUILD THE THREE MAPS
# ---------------------------------------------------------

# (a) Senior Population Map 
p1 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = senior_pop), color = NA) +
  scale_fill_gradient(
    low = "#f2e5ff", high = "#4b0082", 
    name = "Seniors (Count)", na.value = "grey90"
  ) +
 
  geom_text(data = star_locs,
            aes(long, lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +
  
  coord_fixed(1.3) +
  labs(title = "Figure 7a: Total Senior Population (65+) by County") +
  theme_void() + 
  theme(legend.position = "bottom")

# (b) Senior Population Density Map (OUTLIER REMOVED + km²)
# Cap extreme values (stronger than 99%)
upper_limit <- quantile(map_data_all$senior_density, 0.95, na.rm = TRUE)

map_data_all <- map_data_all %>%
  mutate(senior_density_cap = pmin(senior_density, upper_limit))

# (b) Senior Population Density Map (OUTLIER REMOVED + km²)
p2 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = senior_density), color = NA) +
  scale_fill_gradient(
    low = "#e5f5f9", high = "#006d2c", 
    name = "Seniors / km²", na.value = "grey90"
  ) +
  geom_text(data = star_locs,
            aes(long, lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 7b: Senior Population Density (Aged 65+)") +
  theme_void() + 
  theme(legend.position = "bottom")

# (c) % of Senior Population Map
p3 <- ggplot(map_data_all, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pct_senior), color = NA) +
  scale_fill_gradient(
    low = "#fff7bc", high = "#d95f0e", 
    name = "% Senior", na.value = "grey90"
  ) +
  geom_text(data = star_locs,
            aes(long, lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 7c: Percentage of Senior Population (65+)") +
  theme_void() + 
  theme(legend.position = "bottom")

print(p1)
print(p2)
print(p3)
```

------------------------------------------------------------------------

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Prepare Data
missingness_check <- pop_raw %>%
  slice(-1) %>%
  select(
    `Total Population` = S0103_C01_001E,
    `Senior Population (65+)` = S0103_C02_001E
  ) %>%
  mutate(across(everything(), ~na_if(., "*****"))) %>%
  mutate(across(everything(), as.numeric))

# Total number of observations
n_obs <- nrow(missingness_check)

# Calculate Missingness
na_summary_pop <- missingness_check %>%
  summarise(across(everything(), list(
    count = ~sum(is.na(.)),
    pct = ~round(mean(is.na(.)) * 100, 2)
  ))) %>%
  
  pivot_longer(
    everything(),
    names_to = c("Variable Name", ".value"),
    names_pattern = "(.*)_(count|pct)"
  ) %>%
  
  rename(`NA Count` = count, `Missing (%)` = pct) %>%
  
  # Add total observations
  mutate(`Total N` = n_obs) %>%
  
  mutate(Status = case_when(
    `Missing (%)` == 0 ~ "Complete",
    `Missing (%)` < 5 ~ "High Integrity",
    `Missing (%)` < 20 ~ "Conditional",
    TRUE ~ "Limited Usage"
  )) %>%
  
  select(`Variable Name`, `Total N`, `NA Count`, `Missing (%)`, Status)

# 3. Render Table
na_summary_pop %>%
  kbl(caption = "Table 4: Completeness Summary for Census Population Variables") %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  ) %>%
  row_spec(0, bold = TRUE, background = "#f2f2f2") %>%
  column_spec(4, bold = TRUE)
```

### 3.2. Economic Effect

To evaluate socioeconomic influences, the proportion of households receiving Supplemental Nutrition Assistance Program (SNAP) benefits was calculated and mapped at the county level. Strikingly, the spatial distribution of SNAP participation closely mirrored the two curved patterns observed in AD prevalence.

To quantify this relationship, county-level SNAP proportions were plotted against AD prevalence, yielding an $R^2$ value of 0.305. Although moderate, this association supports a potential relationship between socioeconomic disadvantage and AD prevalence. These findings suggest that poverty-related factors may contribute to regional variation in Alzheimer’s disease burden.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Load SNAP data

snap_raw <- read_csv("SNAP.csv", show_col_types = FALSE)

# Remove second row (descriptions)
snap <- snap_raw[-1, ]


# Clean SNAP data

snap_clean <- snap %>%
  select(NAME, S2201_C01_001E, S2201_C03_001E) %>%
  rename(
    location = NAME,
    total_hh = S2201_C01_001E,
    snap_hh  = S2201_C03_001E
  ) %>%
  mutate(
    total_hh = as.numeric(total_hh),
    snap_hh  = as.numeric(snap_hh)
  ) %>%
  separate(location, into = c("county", "state"), sep = ", ") %>%
  mutate(
    county = str_remove(county, " County"),
    county_lower = tolower(county),
    state_lower  = tolower(state),
    pct_snap = (snap_hh / total_hh) * 100
  ) %>%
  select(county_lower, state_lower, pct_snap)


# Map data merge

us_counties <- map_data("county")

map_snap <- us_counties %>%
  left_join(snap_clean,
            by = c("subregion" = "county_lower",
                   "region" = "state_lower"))


#  SNAP % map

p_map <- ggplot(map_snap, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = pct_snap), color = NA) +
  
  scale_fill_gradient(
    low = "#e5ffe5",
    high = "#006400",
    name = "% SNAP HH",
    na.value = "grey90"
  ) +
  geom_text(data = star_locs,
            aes(long, lat, label = "*"),
            color = "yellow",
            size = 6,
            inherit.aes = FALSE) +
  coord_fixed(1.3) +
  labs(title = "Figure 8: County-level % of Households Receiving SNAP (2016-2020)") +
  theme_void() +
  theme(legend.position = "right")
  #theme(
   # plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  #  legend.title = element_text(size = 10),
   # legend.text = element_text(size = 8)
  #)
p_map
```

```{r, echo=FALSE}
library(dplyr)

# Prepare data
scatter_data <- snap_clean %>%
  left_join(ad_prevalence,
            by = c("county_lower", "state_lower")) %>%
  mutate(
    prevalence = as.numeric(prevalence),
    label = paste0(
      "County: ", county_lower,
      "<br>State: ", state_lower,
      "<br>% SNAP: ", round(pct_snap, 2),
      "<br>AD Prevalence: ", round(prevalence, 2)
    )
  ) %>%
  filter(!is.na(pct_snap), !is.na(prevalence))

# =========================
# Fit linear model manually
# =========================
model <- lm(prevalence ~ pct_snap, data = scatter_data)

# Create prediction line
x_seq <- seq(min(scatter_data$pct_snap),
             max(scatter_data$pct_snap),
             length.out = 100)

line_data <- data.frame(
  pct_snap = x_seq,
  prevalence = predict(model, newdata = data.frame(pct_snap = x_seq))
)

# Correlation
cor_val <- cor(scatter_data$pct_snap,
               scatter_data$prevalence,
               use = "complete.obs")

# =========================
# Build plot
# =========================
p <- ggplot(scatter_data,
            aes(x = pct_snap, y = prevalence, text = label)) +
  geom_point(alpha = 0.6) +
  
  # MANUAL regression line (works in plotly)
  geom_line(data = line_data,
            aes(x = pct_snap, y = prevalence),
            color = "blue",
            linewidth = 1,
            inherit.aes = FALSE) +
  
  labs(
    x = "% SNAP Households",
    y = "AD Prevalence (%)",
    title = "% SNAP Households vs Alzheimer's Prevalence"
  ) +
  theme_minimal()

# Convert to interactive + annotation
ggplotly(p, tooltip = "text") %>%
  layout(
    annotations = list(
      x = max(scatter_data$pct_snap),
      y = max(scatter_data$prevalence),
      text = paste0("r = ", round(cor_val, 3)),
      showarrow = FALSE,
      xanchor = "right",
      yanchor = "top"
    )
  )

```
